---
title: "Correlation wheel plots with circlecorR"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Correlation wheel plots with circlecorR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 7,
  dpi = 110,
  fig.align = "center"
)
```

```{r setup}
library(circlecorR)
```

## Motivation

A traditional correlation matrix carries a large amount of **redundant
information**. For $k$ variables it has $k^2$ cells, but only
$k(k-1)/2$ of them are unique: the diagonal is all self-correlations
($r = 1$), the two triangles mirror each other, and large blocks of
*within-category* correlations (symptom--symptom, metric--metric) are often not
the question being asked. As $k$ grows, the interesting **between-category**
relationships are buried in this redundancy and the figure becomes unreadable.

The **correlation wheel** was introduced by Gharibans and colleagues to cut
through this: variables sit around a circle, grouped by category, and only the
correlations of interest are drawn as curved links coloured by their coefficient
[@gharibans2019]. `circlecorR` reproduces that figure natively in R and makes
the grouping, colours, and statistics configurable.

## The simplest path: straight from your data

Most datasets have **one row per subject** and **one column per variable**. You
do not need to build correlation matrices yourself -- hand that data frame
directly to `corr_wheel()` and it computes the correlations (and p-values) for
you.

The only other thing you supply is `groups`: a named list mapping each category
to its variables. This both **selects** which variables appear (any other
columns, such as an ID, are ignored) and sets their **order** around the wheel.

```{r groups}
groups <- list(
  Demographics = c("Age", "BMI"),
  Metrics      = c("Amplitude", "Fed-Fasted AR", "Frequency", "GA-RI"),
  Symptoms     = c("Nausea", "Early satiety", "Bloating",
                   "Upper GI pain", "Lower GI pain", "Heartburn"),
  Scores       = c("GCSI", "PAGI-SYM", "PAGI-QoL", "EQ-5D")
)
```

```{r raw-data}
# `gastro_symptoms` is a synthetic per-row example dataset shipped with the package
head(gastro_symptoms[, 1:5])

corr_wheel(
  gastro_symptoms,             # raw data: one row per subject
  groups      = groups,
  method      = "pearson",     # correlation method
  adjust      = "hochberg",    # multiple-comparison adjustment (see below)
  sig_level   = 0.05,          # hide links with adjusted p > 0.05
  r_threshold = 0.3,           # ...and links with |r| < 0.3
  r_limits    = c(-0.6, 0.6)
)
```

## Hiding self- and within-category correlations

Two of the wheel's most important features are that it **never draws
self-correlations** (the diagonal) and, by default, **hides within-category
correlations** (`hide_within_group = TRUE`). This removes exactly the redundant
parts of the matrix and leaves the between-category structure.

Crucially, this is carried through to the **statistics**. Multiple-comparison
correction penalises you for the number of hypotheses tested. If self- and
within-category correlations are never tested, they should not count towards
that family. `corr_wheel()` therefore applies the adjustment over **only the
correlations it displays**. Shrinking the family makes the correction less
severe -- so power improves -- while remaining statistically consistent with
what is shown.

```{r family}
res <- corr_wheel(gastro_symptoms, groups = groups, adjust = "hochberg",
                  r_threshold = 0.3, r_limits = c(-0.6, 0.6))

k <- length(unlist(groups))
cat("Unique correlations in the full matrix:", k * (k - 1) / 2, "\n")
cat("Correlations actually tested (the family):", res$n_tests, "\n")
```

Because the family is smaller, each raw p-value is corrected by a smaller
factor. Taking Bonferroni for a transparent example, the *same* correlation is
penalised by the number of tests in its family -- here `r res$n_tests` rather
than `r k * (k - 1) / 2`:

```{r power}
p_raw <- 5e-4                       # a raw p-value for one correlation
k_all <- k * (k - 1) / 2

cat("Bonferroni across the full matrix:", signif(p_raw * k_all, 3), "\n")
cat("Bonferroni across the family only:", signif(p_raw * res$n_tests, 3), "\n")
```

The step-up methods (`"holm"`, `"hochberg"`, `"BH"`) behave the same way: fewer
comparisons never gives a larger adjusted p-value, so hiding redundant
self- and within-category correlations can only help power.

If you pass a **pre-computed p-value matrix** instead of raw data, it is taken as
given (no re-adjustment), since it may already be corrected.

## Other inputs

`corr_wheel()` auto-detects what you pass:

* **raw data** (non-square) -- correlations computed for you;
* a **correlation matrix** plus a matching `p` matrix;
* a **`circlecor`** object from `compute_correlations()`.

```{r matrices, eval = FALSE}
r <- as.matrix(read.csv("Rvalues.csv", row.names = 1))
p <- as.matrix(read.csv("Pvalues.csv", row.names = 1))
corr_wheel(r, p, groups = groups, r_threshold = 0.3)
```

## Customising the look

### Colours and labels

`colors` maps categories to colours; `labels` gives pretty display names. Both
are named vectors -- specify only the ones you want to change.

```{r colours}
corr_wheel(
  gastro_symptoms, groups = groups, r_threshold = 0.3, r_limits = c(-0.6, 0.6),
  colors = c(Demographics = "#4C72B0", Metrics = "#DD8452",
             Symptoms = "#55A868", Scores = "#C44E52"),
  labels = c(`GA-RI` = "Rhythm index")
)
```

### Size of blocks and lines

* `tile_height` -- radial thickness of the category **blocks** (smaller = thinner).
* `link_lwd` -- **line** width of the links (larger = thicker).

```{r sizes}
corr_wheel(
  gastro_symptoms, groups = groups, r_threshold = 0.3, r_limits = c(-0.6, 0.6),
  tile_height = 0.12,   # thicker blocks
  link_lwd    = 3       # thicker lines
)
```

### The diverging colour scale

`palette` sets the three colours at `c(-limit, 0, +limit)` and `r_limits` the
scale range.

```{r palette}
corr_wheel(
  gastro_symptoms, groups = groups, r_threshold = 0.3,
  palette  = c("#2166AC", "white", "#B2182B"),   # blue - white - red
  r_limits = c(-0.5, 0.5)
)
```

## Saving to a file

`corr_wheel()` draws on the active graphics device, so save it the usual way:

```{r save, eval = FALSE}
png("correlation_wheel.png", width = 2500, height = 2000, res = 300)
corr_wheel(gastro_symptoms, groups = groups, r_threshold = 0.3,
           r_limits = c(-0.6, 0.6))
dev.off()
```

## References
