---
title: "Correlation wheel plots with circlecorR"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Correlation wheel plots with circlecorR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 7,
  dpi = 110,
  fig.align = "center"
)
```

```{r setup}
library(circlecorR)
```

## Motivation

A traditional correlation matrix carries a large amount of **redundant
information**. For $k$ variables it has $k^2$ cells, but only
$k(k-1)/2$ of them are unique: the diagonal is all self-correlations
($r = 1$), the two triangles mirror each other, and large blocks of
*within-category* correlations (symptom--symptom, metric--metric) are often not
the question being asked. As $k$ grows, the interesting **between-category**
relationships are buried in this redundancy and the figure becomes unreadable.

The **correlation wheel** was introduced by Gharibans and colleagues to cut
through this: variables sit around a circle, grouped by category, and only the
correlations of interest are drawn as curved links coloured by their coefficient
[@gharibans2019]. `circlecorR` reproduces that figure natively in R and makes
the grouping, colours, and statistics configurable.

## Installation

```r
# install.packages("remotes")
remotes::install_github("kriz98/circlecorR", build_vignettes = TRUE)
```

*(CRAN release pending.)*

Dependencies (`circlize` and `psych`) install automatically with the command
above.

## Quick start: straight from your data

`circlecorR` is designed to work straight from a data frame -- **one row per
subject** and **one column per variable** -- with no separate step to build
correlation matrices yourself. Hand that data frame directly to `corr_wheel()`
and it computes the correlations (and p-values) for you.

The only other thing you supply is `groups`: a named list mapping each category
to its variables. This both **selects** which variables appear (any other
columns, such as an ID, are ignored) and sets their **order** around the wheel.

`gastro_symptoms`, used throughout this vignette, is a synthetic example
dataset bundled with the package. Because the package sets `LazyData: true`,
it's available directly as soon as you `library(circlecorR)` -- no `data()`
call needed. Swap it for your own data frame (one row per subject) to use your
own data; see `?gastro_symptoms` for its exact shape.

```{r groups}
groups <- list(
  Demographics = c("Age", "BMI"),
  Metrics      = c("Amplitude", "Fed-Fasted AR", "Frequency", "GA-RI"),
  Symptoms     = c("Nausea", "Early satiety", "Bloating",
                   "Upper GI pain", "Lower GI pain", "Heartburn"),
  Scores       = c("GCSI", "PAGI-SYM", "PAGI-QoL", "EQ-5D")
)
```

```{r raw-data}
# `gastro_symptoms` is a synthetic per-row example dataset shipped with the package
head(gastro_symptoms[, 1:5])

corr_wheel(
  gastro_symptoms,             # raw data: one row per subject
  groups      = groups,
  method      = "pearson",     # correlation method
  adjust      = "hochberg",    # multiple-comparison adjustment (see below)
  sig_level   = 0.05,          # hide links with adjusted p > 0.05
  r_threshold = 0.3,           # ...and links with |r| < 0.3
  r_limits    = c(-0.6, 0.6)
)
```

## Hiding self- and within-category correlations

Two of the wheel's most important features are that it **never draws
self-correlations** (the diagonal) and, by default, **hides within-category
correlations** (`hide_within_group = TRUE`). This removes exactly the redundant
parts of the matrix and leaves the between-category structure.

Crucially, this is carried through to the **statistics**. Multiple-comparison
correction penalises you for the number of hypotheses tested. If self- and
within-category correlations are never tested, they should not count towards
that family. `corr_wheel()` therefore applies the adjustment over **only the
correlations it displays**. Shrinking the family makes the correction less
severe -- so power improves -- while remaining statistically consistent with
what is shown.

```{r family}
res <- corr_wheel(gastro_symptoms, groups = groups, adjust = "hochberg",
                  r_threshold = 0.3, r_limits = c(-0.6, 0.6))

k <- length(unlist(groups))
cat("Unique correlations in the full matrix:", k * (k - 1) / 2, "\n")
cat("Correlations actually tested (the family):", res$n_tests, "\n")
```

Because the family is smaller, each raw p-value is corrected by a smaller
factor. Taking Bonferroni for a transparent example, the *same* correlation is
penalised by the number of tests in its family -- here `r res$n_tests` rather
than `r k * (k - 1) / 2`:

```{r power}
p_raw <- 5e-4                       # a raw p-value for one correlation
k_all <- k * (k - 1) / 2

cat("Bonferroni across the full matrix:", signif(p_raw * k_all, 3), "\n")
cat("Bonferroni across the family only:", signif(p_raw * res$n_tests, 3), "\n")
```

The step-up methods (`"holm"`, `"hochberg"`, `"BH"`) behave the same way: fewer
comparisons never gives a larger adjusted p-value, so hiding redundant
self- and within-category correlations can only help power.

If you want the correlation and p-value matrices themselves -- for a table, a
report, or any other purpose -- `compute_correlations()` is the same function
`corr_wheel()` calls internally, available on its own:

```{r compute-correlations}
cc <- compute_correlations(gastro_symptoms, method = "pearson")
str(cc)
```

## Customising the look

### Colour schemes

`scheme` sets the category colours and the diverging link palette together, as
one named preset. Built-in options are listed by `corr_wheel_schemes()`:

```{r schemes-list}
corr_wheel_schemes()
```

* `"default"` -- the seaborn-like categorical palette and a blue-white-red
  diverging scale used throughout this vignette.
* `"colorblind"` -- the Okabe-Ito categorical palette with a colourblind-safe
  (PuOr) diverging scale.
* `"ocean"` -- hue-varied cool tones (teal, blue, indigo, violet, slate) for
  categories, with a blue-to-warm diverging scale.
* `"vivid"` -- a brighter, higher-contrast categorical palette, still paired
  with the blue-white-red diverging scale.
* `"alimetry"` -- blues alternating with warm golds, with a matching
  blue-to-yellow diverging scale.

```{r scheme-colorblind}
corr_wheel(gastro_symptoms, groups = groups, r_threshold = 0.3,
          r_limits = c(-0.6, 0.6), scheme = "colorblind")
```

```{r scheme-alimetry}
corr_wheel(gastro_symptoms, groups = groups, r_threshold = 0.3,
          r_limits = c(-0.6, 0.6), scheme = "alimetry")
```

`corr_wheel_scheme()` returns a scheme's definition so you can start from a
preset and tweak it -- here, lightening the midpoint of the diverging scale:

```{r scheme-tweak}
s <- corr_wheel_scheme("ocean")
s$palette[2] <- "grey96"
corr_wheel(gastro_symptoms, groups = groups, r_threshold = 0.3,
          r_limits = c(-0.6, 0.6), scheme = s)
```

Or build one entirely from scratch by passing `list(colors = , palette = )`
directly -- `colors` is an unnamed vector cycled across however many
categories you have, and `palette` is the length-3 diverging scale (negative,
midpoint, positive):

```{r scheme-custom}
corr_wheel(gastro_symptoms, groups = groups, r_threshold = 0.3,
          r_limits = c(-0.6, 0.6),
          scheme = list(colors = c("#7B2CBF", "#2A9D8F", "#E76F51", "#264653"),
                       palette = c("#2A9D8F", "white", "#E76F51")))
```

### Colours and labels

`colors` maps categories to colours, layered on top of `scheme` (or the
default palette if `scheme` is `NULL`) -- only the categories you name are
overridden. `labels` gives pretty display names. Both are named vectors --
specify only the ones you want to change.

```{r colours}
corr_wheel(
  gastro_symptoms, groups = groups, r_threshold = 0.3, r_limits = c(-0.6, 0.6),
  scheme = "colorblind",
  colors = c(Scores = "black"),          # override just one category
  labels = c("GA-RI" = "Rhythm index")
)
```

### Size of blocks and lines

* `tile_height` -- radial thickness of the category **blocks** (smaller = thinner).
* `link_lwd` -- **line** width of the links (larger = thicker).

```{r sizes}
corr_wheel(
  gastro_symptoms, groups = groups, r_threshold = 0.3, r_limits = c(-0.6, 0.6),
  tile_height = 0.12,   # thicker blocks
  link_lwd    = 3       # thicker lines
)
```

### The diverging colour scale

`palette` sets the three colours at `c(-limit, 0, +limit)` (overriding
`scheme`'s) and `r_limits` the scale range.

```{r palette}
corr_wheel(
  gastro_symptoms, groups = groups, r_threshold = 0.3,
  palette  = c("#2166AC", "white", "#B2182B"),   # blue - white - red
  r_limits = c(-0.5, 0.5)
)
```

## Saving to a file

`corr_wheel()` draws on the active graphics device, so save it the usual way:

```{r save, eval = FALSE}
png("correlation_wheel.png", width = 2500, height = 2000, res = 300)
corr_wheel(gastro_symptoms, groups = groups, r_threshold = 0.3,
           r_limits = c(-0.6, 0.6))
dev.off()
```

## Argument reference

Everything covered above, at a glance:

| What | Argument | Example |
|------|----------|---------|
| Category assignment & order | `groups` | named list *or* `variable = category` vector |
| Colour scheme (category colours + link palette, together) | `scheme` | `"colorblind"`, `"ocean"`, `"vivid"`, `"alimetry"`, or `list(colors=, palette=)` |
| Category colours | `colors` | `c(Symptoms = "#55A868", Scores = "#C44E52")` (overrides `scheme` per category) |
| Pretty variable labels | `labels` | `c("GA-RI" = "Rhythm index")` |
| Significance cutoff | `sig_level` | `0.05` |
| Multiple-comparison adjustment | `adjust` | `"holm"`, `"hochberg"`, `"BH"`, `"none"` |
| Minimum \|r\| shown | `r_threshold` | `0.3` |
| Hide within-category links | `hide_within_group` | `TRUE` / `FALSE` |
| Colour-scale range | `r_limits` | `c(-0.5, 0.5)` |
| Link colour ramp | `palette` | `c("#2166AC", "white", "#B2182B")` (overrides `scheme`'s) |
| Block size (thickness) | `tile_height` | `0.06` (thin) … `0.12` (thick) |
| Line size (width) | `link_lwd` | `1.6`, `3` |
| Rotation / spacing | `start_degree`, `group_gap`, `node_gap` | |
| Legend / colour bar | `legend`, `colorbar` | |

`corr_wheel()` returns (invisibly) the ordered variables, resolved group and
colour maps, the colour function, and the masked matrix actually plotted --
handy for reproducibility or building a caption.

## References
